---
title: "Codebook"
author: "Duncan Turnbull"
date: "Sunday, December 21, 2014"
output: pdf_document
---

#CodeBook

Raw data activity measurements have been transformed into summaries in a tidy data format. The summary is IDs into Names

## Data file
The data file is: tidyoutput.txt

It contains a header row defining the variables - explanations below
It is 180 rows by 68 columns. The first two columns are the grouping columns: ActivityName and Subject
The next 66 columns are the variables summarised by averagig for each subject by activity type. 

e.g. (truncated for clarity)

```R
> head(tidyoutput[1:7],2)
Source: local data frame [2 x 7]
Groups: ActivityName

  ActivityName Subject tBodyAccmeanX tBodyAccmeanY tBodyAccmeanZ tBodyAccstdX tBodyAccstdY
1       LAYING       1     0.2215982   -0.04051395    -0.1132036   -0.9280565   -0.8368274
2       LAYING       2     0.2813734   -0.01815874    -0.1072456   -0.9740595   -0.9802774

```

The activity subject is doing in observation 1 is LAYING and the subject id is 1
The tBodyAccmeanX value of 0.2215982 is the average of all the observations for subject 1 while LAYING.

##Explanations 
**ActivityName**: These are the names of the activities the subjects were completing during the observation (hopefully self explanatory). The values are:
    
    - WALKING
    - WALKING_UPSTAIRS
    - WALKING_DOWNSTAIRS
    - SITTING
    - STANDING
    - LAYING

**Subject**: This is the numeric id of the subject from whom the measurements were taken. Value range from 1 to 30

Every other variable in the output is an **average** of the measurements for the specific subject and  the activity they were doing. 

The variable names are made up of representative info as follows:

* These base measurements observed (before summarisation) were normalised and bounded within [-1,1].
* The observation summary is either time or frequency. This is represented with the first letter: t or f 
* The end of the name will be std or mean then X,Y,Z or blank. 
    * The X Y or Z letter represents the direction the signal reported.
    * The mean or std represents whether this summarised observations were means or standard deviations. 
* The remaining label between these values represent the type of sensor, body or gyroscope, acceleration e.g. tGravityAccstdZ will show the average of the standard deviation of a time variable of Gravity Acceleration in the Z direction for the respective subject carrying out the respective activity

The columns are broken out for clarity a little more in this table:

```{r eval=FALSE}
> codebook
                      codes timefreq measure direction             feature
1              ActivityName                                   ActivityName
2                   Subject                                        Subject
3             tBodyAccmeanX        t    mean         x             BodyAcc
4             tBodyAccmeanY        t    mean         y             BodyAcc
5             tBodyAccmeanZ        t    mean         z             BodyAcc
6              tBodyAccstdX        t     std         x             BodyAcc
7              tBodyAccstdY        t     std         y             BodyAcc
8              tBodyAccstdZ        t     std         z             BodyAcc
9          tGravityAccmeanX        t    mean         x          GravityAcc
10         tGravityAccmeanY        t    mean         y          GravityAcc
11         tGravityAccmeanZ        t    mean         z          GravityAcc
12          tGravityAccstdX        t     std         x          GravityAcc
13          tGravityAccstdY        t     std         y          GravityAcc
14          tGravityAccstdZ        t     std         z          GravityAcc
15        tBodyAccJerkmeanX        t    mean         x         BodyAccJerk
16        tBodyAccJerkmeanY        t    mean         y         BodyAccJerk
17        tBodyAccJerkmeanZ        t    mean         z         BodyAccJerk
18         tBodyAccJerkstdX        t     std         x         BodyAccJerk
19         tBodyAccJerkstdY        t     std         y         BodyAccJerk
20         tBodyAccJerkstdZ        t     std         z         BodyAccJerk
21           tBodyGyromeanX        t    mean         x            BodyGyro
22           tBodyGyromeanY        t    mean         y            BodyGyro
23           tBodyGyromeanZ        t    mean         z            BodyGyro
24            tBodyGyrostdX        t     std         x            BodyGyro
25            tBodyGyrostdY        t     std         y            BodyGyro
26            tBodyGyrostdZ        t     std         z            BodyGyro
27       tBodyGyroJerkmeanX        t    mean         x        BodyGyroJerk
28       tBodyGyroJerkmeanY        t    mean         y        BodyGyroJerk
29       tBodyGyroJerkmeanZ        t    mean         z        BodyGyroJerk
30        tBodyGyroJerkstdX        t     std         x        BodyGyroJerk
31        tBodyGyroJerkstdY        t     std         y        BodyGyroJerk
32        tBodyGyroJerkstdZ        t     std         z        BodyGyroJerk
33          tBodyAccMagmean        t    mean                    BodyAccMag
34           tBodyAccMagstd        t     std                    BodyAccMag
35       tGravityAccMagmean        t    mean                 GravityAccMag
36        tGravityAccMagstd        t     std                 GravityAccMag
37      tBodyAccJerkMagmean        t    mean                BodyAccJerkMag
38       tBodyAccJerkMagstd        t     std                BodyAccJerkMag
39         tBodyGyroMagmean        t    mean                   BodyGyroMag
40          tBodyGyroMagstd        t     std                   BodyGyroMag
41     tBodyGyroJerkMagmean        t    mean               BodyGyroJerkMag
42      tBodyGyroJerkMagstd        t     std               BodyGyroJerkMag
43            fBodyAccmeanX        f    mean         x             BodyAcc
44            fBodyAccmeanY        f    mean         y             BodyAcc
45            fBodyAccmeanZ        f    mean         z             BodyAcc
46             fBodyAccstdX        f     std         x             BodyAcc
47             fBodyAccstdY        f     std         y             BodyAcc
48             fBodyAccstdZ        f     std         z             BodyAcc
49        fBodyAccJerkmeanX        f    mean         x         BodyAccJerk
50        fBodyAccJerkmeanY        f    mean         y         BodyAccJerk
51        fBodyAccJerkmeanZ        f    mean         z         BodyAccJerk
52         fBodyAccJerkstdX        f     std         x         BodyAccJerk
53         fBodyAccJerkstdY        f     std         y         BodyAccJerk
54         fBodyAccJerkstdZ        f     std         z         BodyAccJerk
55           fBodyGyromeanX        f    mean         x            BodyGyro
56           fBodyGyromeanY        f    mean         y            BodyGyro
57           fBodyGyromeanZ        f    mean         z            BodyGyro
58            fBodyGyrostdX        f     std         x            BodyGyro
59            fBodyGyrostdY        f     std         y            BodyGyro
60            fBodyGyrostdZ        f     std         z            BodyGyro
61          fBodyAccMagmean        f    mean                    BodyAccMag
62           fBodyAccMagstd        f     std                    BodyAccMag
63  fBodyBodyAccJerkMagmean        f    mean            BodyBodyAccJerkMag
64   fBodyBodyAccJerkMagstd        f     std            BodyBodyAccJerkMag
65     fBodyBodyGyroMagmean        f    mean               BodyBodyGyroMag
66      fBodyBodyGyroMagstd        f     std               BodyBodyGyroMag
67 fBodyBodyGyroJerkMagmean        f    mean           BodyBodyGyroJerkMag
68  fBodyBodyGyroJerkMagstd        f     std           BodyBodyGyroJerkMag
```

This is the code for creating that table. It just looks and changes particular values to things a little more easy to understand.


```{r eval=FALSE}
codebook <- data.frame(codes,timefreq= "", measure ="",direction = "", stringsAsFactors = FALSE)
codebook$timefreq[c(grep(codebook$codes, pattern = "^t"))] <- "t"
codebook$timefreq[c(grep(codebook$codes, pattern = "^f"))] <- "f"
codebook$direction[c(grep(codebook$codes, pattern = "X$"))] <- "x"
codebook$direction[c(grep(codebook$codes, pattern = "Y$"))] <- "y"
codebook$direction[c(grep(codebook$codes, pattern = "Z$"))] <- "z"
codebook$measure[c(grep(codebook$codes, pattern = "mean"))] <- "mean"
codebook$measure[c(grep(codebook$codes, pattern = "std"))] <- "std"
codebook$feature <- gsub("^[tf](.*)(mean|std)|","\\1",codebook$codes)
codebook$feature <- gsub( pattern = "(.*)[XYZ]$","\\1" ,codebook$feature)
write.table( codebook, row.name=FALSE ,file = "./codebook.txt")
```